The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
6.2.3.

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.z

\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

>>>>

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}
a
We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interace \citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.
